---
title: "Lab 06 - Ugly charts and Simpson's paradox"
author: "Marq Schieber"
date: "3/8/2022"
output: github_document
---

### Load packages and data

```{r load-packages, message = FALSE}
library(tidyverse) 
library(mosaicData)
```

### Exercise 1

```{r}
staff <- read_csv("data/instructional-staff.csv")
```

How many rows will the long-format data have? It will have a row for each combination of year and faculty type. If there are 5 faculty types and 11 years of data, how many rows will we have?

Will there be 55 rows?

```{r}
staff_long <- staff %>%
  pivot_longer(cols = -faculty_type, names_to = "year") %>%
  mutate(value = as.numeric(value))
```

```{r line plot 1}
staff_long %>%
  ggplot(aes(x = year, y = value, group = faculty_type, color = faculty_type)) +
  geom_line()+
  labs(title = "Trends in Instructional Staff Employment Status, 1975-2011", y = "Percent", x = "Year", color = "Faculty Type")

```

Suppose the objective of this plot was to show that the proportion of part-time faculty have gone up over time compared to other instructional staff types. > What changes would you propose making to this plot to tell this story?

I would make the graph a series of collumns, with faculty_type as the fill.

```{r}
staff_long %>% 
  ggplot(aes(x = year, y = value , fill = faculty_type)) +
  geom_col()
```


### Exercise 2

Can you help them make improve it? First, brainstorm how you would improve it. Then create the improved visualization and write up the changes/decisions you made as bullet points. It’s ok if some of your improvements are aspirational, i.e. you don’t know how to implement it, but you think it’s a good idea. Ask a tutor for help, but also keep an eye on the time. Implement what you can and leave note identifying the aspirational improvements.

I think you would first want to limit which countries as listed on the graph. Perhaps, those countries that have a total fish of at least 600,000. I think a bar graph that has total count on the Y axis and th fill consisitng of capture and aquaculture fish would be useful visually. Furthermore, you could order countries either from greatest to least or in terms of another measure (i.e. population). It depends on what you're trying to show with this presetned information.

```{r}
fisheries <- read_csv("data/fisheries.csv")

```

#got some help from Elayna, but need to ask her again. I think I need to specifiy what values go into harvest.
```{r}
filtered_fisheries<-fisheries %>% 
filter(total>600000)

filtered_fisheries <- filtered_fisheries %>%
  pivot_longer(cols = -country, names_to = "harvest") %>%
  mutate(value = as.numeric(value))

filtered_fisheries %>% 
  ggplot(aes(x = country, y = value, fill = harvest)) +
  geom_col()

```

### Exercise 3

What type of study do you think these data come from: observational or experiment? Why?

Observational. Effect of smoking on lifespan. It would be unethical to assign a 'potentially' toxic substance for a longitudinal study.

### Exercise 4
How many observations are in this dataset? What does each observation represent?

1314 obsercations. Each observation is a person and whether they're still alive or not.

### Exercise 5
How many variables are in this dataset? What type of variable is each? Display each variable using an appropriate visualization.

3 variables. Outcome and smoker are nominal variables. Age is an ordinal variable.

### Exercise 6
What would you expect the relationship between smoking status and health outcome to be?

Smokers live shorter lives than non smokers. Higher mortality rate too probably.

### Exercise 7
Create a visualization depicting the relationship between smoking status and health outcome. Briefly describe the relationship, and evaluate whether this meets your expectations. Additionally, calculate the relevant conditional probabilities to help your narrative. Here is some code to get you started:

There does not appear to be a relationship between smoking status and health outcome...Although I think that would change if we were to look at age of "outcome" aka death. 

```{r}
Whickham %>%
  count(smoker, outcome) -> count

count %>%
ggplot(aes(x = smoker, y = n, fill = outcome)) +
  geom_bar(stat="identity", position=position_dodge())

Whickham %>% 
  count(smoker, outcome) %>% 
  group_by(smoker) %>% 
  mutate(Percentage = n / sum(n))

```


### Exercise 8
Create a new variable called age_cat using the following scheme:
age <= 44 ~ "18-44"
age > 44 & age <= 64 ~ "45-64"
age > 64 ~ "65+"

```{r new variable}
Whickham <- Whickham %>% 
  mutate(age_cat = case_when(
    age <= 44 ~"18-44",
    age > 44 & age <= 64 ~ "44-64",
    age > 64 ~ "65+"))
```


### Exercise 9
Re-create the visualization depicting the relationship between smoking status and health outcome, faceted by age_cat. What changed? What might explain this change? Extend the contingency table from earlier by breaking it down by age category and use it to help your narrative.

```{r}
Whickham %>%
  count(smoker, age_cat, outcome)

Whickham %>% 
  ggplot(aes(x = smoker, fill = outcome)) +
  geom_bar()+
  facet_wrap(~age_cat)

Whickham %>% 
  count(smoker, outcome, age_cat) %>% 
  group_by(smoker, age_cat) %>% 
  mutate(Percentage = n / sum(n))

```

Now when  you look at the effect of smoking with age....People who smoke die at younger ages than those who don't smoke. This suggests that smoking decreases people's lifespans. 
